Towards an Automated Sportscasting System for Chess

### 2011-05-18 paper/paper.tex
added svmlight cite

\begin{abstract}
We investigate applications of statistical learning to automated sportscasting for chess. In particular, we 
\end{abstract}

\section{Introduction}
Chess has a long history of intriguing enthusiasts due to its complexity. Even after years of play, each game is new challenge, a new problem to solve. Historically, chess has been prominent as a sport where the battles of grandmasters engulf thousands; given the highly intricate nature of chess (and similar abstract strategy games), experts have scrutinized these games to devise a standard high-level structures like openings, strategies, tactics, and time advantage, which they use to explain maneuvers and blunders that occur in the games. Devotees of chess also attempt to improve their own performance in terms of these high-level structures, reading chess theorists' comments about games by high-ranking players.

@@

\paragraph{Automated sportscasting.} Chen and Mooney (\cite{chen}) present a commentator system that learns language from sportscasts of simulated RoboCup games. Their system, which is trained on ambiguous supervision in the form of textual human commentaries and simulation states of the soccer games, simultaneously tries to establish correspondences between the commentaries and the simulation states. Their work focuses on generating meaningful English commentary from relatively simple topics such as ‘Player 1 kicked the ball to Player 2’, while our work focuses more on identifying complex topics.

\paragraph{Automatic chess commentary.} There exist techniques for automatic annotation of chess games which are heavily dependent on rule-based logic and use of a chess-engine to determine which parts of a specific chess game are salient~\cite{cambridge-chess-annotation}. As a departure from the rule-based approach to automatic chess commentary, Guid et al. (\cite{Guid}) discuss the use of argument based machine learning, which analyzes reasoning where arguments for and against a certain claim are produced and evaluated (in the context of chess, why a certain move was played over others.) Their system, much like the one presented in this paper, identifies complex positional features of a given chess game.

@@

\subsection{Parsing}
We tried various open source PGN parsers, each succumbing to flaws like mishandling uncommon cases (such as  promotions and special moves like \emph{en passant}). We based our system on Chesspresso,\footnote{url{http://www.chesspresso.org/}} a Java library for writing chess programs, which we fixed and modified for our needs. Chesspresso takes a PGN file as input and returns a data structure that represents the initial position of the board and the subsequent moves (as well as all metadata necessary to recreate the PGN file). We also extended it to be more robust to variations in input format, in addition to being able to export PGNs in a canonical form, which was crucial for removing duplicates.

\subsection{Preparing the Data}
After parsing each game, we filtered out the tagged games that had more than 10 plies because it will be very difficult to identify the tags that apply to a complex sequence of moves. Among the games that we kept, we also kept only the first few positions and moves (the ``prefix'' of each game, as described earlier), a parameter that can be adjusted and is one by default. We then generate tags for each game based on the information from the source websites. For example, games in a file that corresponds to a lesson called ``Double Attack'' are labeled doubleAttack. In addition, some of those games may appear in other lessons, and consequently will have multiple labels. An additional step that we take during experimentation is to segment the games further by the color of the player in addition to the tags of the game in order to account for the white-bias in the training data.

\section{Features}
\subsection{Sparse auto-encoder}
The sparse auto-encoder is one of many methods in use for unsupervised feature learning~\cite{Coates}. It takes the form of a feedforward neural network, which consists of an input layer of N nodes, some number of hidden layers, and output layer of M nodes. The value of each layer $\vec{l_i}$ is determined with a weight matrix $W_i$ and a bias vector $b_i$: \[L_i = f(W_i \cdot \vec{l_{i-1}} + b_i),\] where $f(\cdot)$ is a non-linearity like the sigmoid $f(x) = \frac{1}{1 + e^{-z}}$ applied component-wise.

@@

A normal auto-encoder only has one layer of hidden nodes, but we can also add more hidden layers to form a stacked auto-encoder. Per~\cite{Bengio07greedylayer-wise}, we use a greedy strategy to train the extra hidden layers in the auto-encoder. We train an auto-encoder with one hidden layer as usual; once this completes, we obtain the output of the hidden layer over all the examples in the input set, and use that as the new set of examples. This allows us to learn higher-level representations of the data~\cite{Vincent}.

@@

\subsection{Manually-generated features}
\begin{table*}
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Category} & \textbf{Description} \\ \hline
Position features & Piece location, potential moves per piece, material count, square control. \\
Move features & Capturing move, recapturing move, en passant move, long castling move, short castling move, check move, stalemate move, checkmate move, promotion move. \\
\hline
\end{tabular} 
\caption{List of chess-specific manual features per position.}
\end{table*}

As a point of comparison to auto-encoders, we selected chess-specific manual features from a given position-move pair.  The position-based features involve information such as where the pieces are on the board and potential moves, while the move-based features involve binary information such as if the move is a capturing move or a check.  We list all our hand-selected manual features in Table REF.

\section{Classification}
We built a support vector machine for structured classification of games. We filtered our tagged games by the number of plies (number of individual moves), generated mappings between tags and game prefixes (some initial position plus a sequence of moves), segmented our tagged games into three groups (training, validation and test), and called SVMLight with that data, an external library for SVM.

\subsection{High-level descriptions}
We classify segments of the chess game based on various high-level descriptions from chess theory and folklore:

\section{Experiments}
\subsection{SVM Based Structured Classifier}
In our experiments, we processed the input data into multimaps from game prefixes to tags. We then segmented the data into three disjoint sets: 90% for training, 5% as the validation set (for tuning our classifiers), and 5% as the test set. The segmenting was done by randomly sampling from the set of all labeled games.

Afterwards, we constructed lists of positive and negative labels (mappings from tags to games that have that tag, and games that do not have that tag). We then built SVM models using the training data. In testing, we fed that to the SVM to generate, for each test game and tag, a boolean prediction of whether that tag applies to that game. As a result, we constructed a SVM-based multiple-class classifier that attempts to predict the tags that characterize an input game.

We measured the accuracy of our model using various measures like false positive rate, false negative rate, precision, recall, and F1. In addition, we compared the predictions made by SVM to those made randomly where each game was assigned a tag at random based on the proportion of the games that have that tag. The random predictions were made 1,000 times, with the predictions made independently at random, and the error rates were averaged.

@@

\begin{table*}
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Category} & \textbf{Description} \\ \hline
Position features & Piece location, potential moves per piece, material count, square control. \\
Move features & Capturing move, recapturing move, en passant move, long castling move, short castling move, check move, stalemate move, checkmate move, promotion move. \\
\hline
\end{tabular} 
\caption{Sources of error in the SVM classifier.}
\label{table:negatives}
\end{table*}

@@

TALK ABOUT USING AUTOENCODER inputs for the SVM

@@

\subsection{Classification Quality}
Future work should emphasize the quality of classifications. We demonstrated that general approaches may not be effective enough to be useful for complex games like chess. Therefore, future projects should seek improvements by better taking advantage of the chess-specific tools and libraries that are available like game engines and weight functions. If the classifier incorporated chess-specific features like those generated by chess engines, it could be much better at identifying the types of strategies that are available to the player. For example, a deep traversal of many future paths by a chess engine may reveal that an attack is possible (attackPossible) or that the player can force a pawn promotion (pawnPromotion). While a purely computational approach to identifying opportunities and strategies may provide higher-quality or more reliable assessments, a statistical approach may be more robust. %talk about what i mean by robust

@@

Another possible improvement is to use a different classifier than SVM, such as the perceptron or maximum entropy classifiers. One issue with the SVM is that often the tags are not so clear-cut: What does it mean for example that an endgame is “good”? Rather, there are degrees of how good an endgame is to the player. Therefore, a softer approach like maximum entropy may produce better results.

\subsection{Generalization to Other Games.}
One goal of this project was to develop a statistical machine learning approach that was general enough to be easily adapted to other applications like StarCraft or Checkers. A future work can explore this possibility by simply creating features for other applications.

>>>>

\begin{abstract}
We investigate applications of statistical learning to automated sportscasting for chess. In particular, this project studies methods to identify high level structures in chess games using support vector machines and autoencoders. 
\end{abstract}

\section{Introduction}
Chess has a long history of intriguing enthusiasts due to its complexity. Even after years of play, each game is new challenge, a new problem to solve. Historically, chess has been prominent as a sport where the battles of grandmasters engulf thousands; given the highly intricate nature of chess (and similar abstract strategy games), experts have scrutinized these games to devise standard high-level structures like openings, strategies, tactics, and time advantage, which they use to explain maneuvers and blunders that occur in the games. Devotees of chess also attempt to improve their own performance in terms of these high-level structures, reading chess theorists' comments about games by high-ranking players.

@@

\paragraph{Automated sportscasting.} Chen and Mooney~\shortcite{chen} present a commentator system that learns language from sportscasts of simulated RoboCup games. Their system, which is trained on ambiguous supervision in the form of textual human commentaries and simulation states of the soccer games, simultaneously tries to establish correspondences between the commentaries and the simulation states. Their work focuses on generating meaningful English commentary from relatively simple topics such as ‘Player 1 kicked the ball to Player 2’, while our work focuses more on identifying complex topics.

\paragraph{Automatic chess commentary.} There exist techniques for automatic annotation of chess games which are heavily dependent on rule-based logic and use of a chess-engine to determine which parts of a specific chess game are salient~\cite{cambridge-chess-annotation}. As a departure from the rule-based approach to automatic chess commentary, Guid et al.~\shortcite{Guid} discuss the use of argument based machine learning, which analyzes reasoning where arguments for and against a certain claim are produced and evaluated (in the context of chess, why a certain move was played over others.) Their system, much like the one presented in this paper, identifies complex positional features of a given chess game.

@@

\subsection{Parsing}
We tried various open source PGN parsers, each succumbing to flaws like mishandling uncommon cases (such as  promotions and special moves like \emph{en passant}). We based our system on Chesspresso,\footnote{\url{http://www.chesspresso.org/}} a Java library for writing chess programs, which we fixed and modified for our needs. Chesspresso takes a PGN file as input and returns a data structure that represents the initial position of the board and the subsequent moves (as well as all metadata necessary to recreate the PGN file). We also extended it to be more robust to variations in input format, in addition to being able to export PGNs in a canonical form, which was crucial for removing duplicates.

\subsection{Preparing the Data}
After parsing each game, we filtered out the tagged games that had more than 10 plies because it will be very difficult to identify the tags that apply to a complex sequence of moves. Among the games that we kept, we also kept only the first few positions and moves (the ``prefix'' of each game, as described earlier), a parameter that can be adjusted; for our experiments, we kept just the initial board specified in each game. We then generate labels for each game based on the information from the source websites. For example, games in a file that corresponds to a lesson called ``Double Attack'' are labeled doubleAttack. In addition, some of those games may appear in other lessons, and consequently will have multiple labels. We also segment the games further by the color of the player that makes the first move in the game, in addition to the tags of the game in order to account for the white-bias in the training data.

\section{Feature Extraction}
\subsection{Sparse auto-encoder}
The sparse auto-encoder is one of many methods in use for unsupervised feature learning~\cite{Coates}. It takes the form of a feedforward neural network, which consists of an input layer of N nodes, some number of hidden layers, and output layer of M nodes. The value of each layer $\vec{l_i}$ is determined with a weight matrix $W_i$ and a bias vector $b_i$: \[L_i = f(W_i \cdot \vec{l_{i-1}} + b_i),\] where $f(\cdot)$ is a non-linearity like the sigmoid $f(x) = \frac{1}{1 + e^{-z}}$ applied component-wise.

@@

A normal auto-encoder only has one layer of hidden nodes, but we can also add more hidden layers to form a stacked auto-encoder. Per Bengio et al.~\shortcite{Bengio07greedylayer-wise}, we use a greedy strategy to train the extra hidden layers in the auto-encoder. We train an auto-encoder with one hidden layer as usual; once this completes, we obtain the output of the hidden layer over all the examples in the input set, and use that as the new set of examples. This allows us to learn higher-level representations of the data~\cite{Vincent}.

@@

\subsection{Manually-generated features}
\begin{table*}
\centering
\begin{tabular}{lp{0.6\textwidth}}
\hline
\textbf{Category} & \textbf{Description} \\ \hline
Position & Piece location, potential moves per piece, material count, square control. \\
Previous move & Capturing move, recapturing move, en passant move, long castling move, short castling move, check move, stalemate move, checkmate move, promotion move. \\
\hline
\end{tabular} 
\caption{List of chess-specific manual features per position.}
\label{table:features}
\end{table*}

As a point of comparison to auto-encoders, we extracted chess-specific features manually from a given chess position (configuration of pieces on the board) and the move that resulted in that position; Table~\ref{table:features} contains a list. The position-based features include information such as where the pieces are on the board and potential moves, while the move-based features involve binary information such as if the move that resulted in this position was a capturing move or resulted in a check.

\section{Experimental methodology}
We processed all of the labeled training data to obtain a mapping from each unique game prefix (for most of the experiments, just the initial specified configuration of the board in the labeled game) to the labels that apply to it; most of the prefixes only had one label, while others had two or three. We then segmented the data into three disjoint sets: 90% for training, 5% as the validation set (for use with tuning our classifiers), and 5% as the test set. The segmenting was done by random selection from the set of all labeled games.

In our experiments, we first use the labeled data in the training set (and possibly also unlabeled data) to train the classifiers. Then, for every test input chess game (or a prefix of a game), we use our classifiers to apply some number of labels to it that come from Table~\ref{table:themes}. We measure the performance of the predictions made by our classifiers by counting the number of true positives (a predicted label was among the set of gold labels for a test input), true negatives (a label in Table~\ref{table:themes} was neither in the set of predicted labels nor the set of gold labels), false positives (a predicted label was not in the set of gold labels), and false negatives (a gold label was not predicted). We compute and report the precision, recall, and F1 score using these measures of accuracy.

\subsection{Support vector machine classifier}
We note that any number of labels may apply to some test input that we want to classify; a straightforward use of a binary (or even multiclass) classifier, as provided by most implementations of support vector machines, will not suffice as it can only assign one label. Therefore, we take the one-versus-all strategy of training many binary SVMs; for each label that we consider, we train a binary classifier with all training games that have the given label as positive examples, and all other games as negative examples. We used SVMlight~\cite{svmlight}, a popular third-party implementation of SVMs, for our experiments, using the manually-generated features extracted from each labeled data point.

\subsection{Classifying with the auto-encoder}
There exists several methods for using the auto-encoder for the task at hand. First, we can consider using the output of the last hidden layer as the feature vector for input into the SVM classifier, just as we did with the manually-generated features. To do this,

\subsection{Random baseline}
In addition, we compared the predictions made by SVM to those made randomly where each game was assigned a tag at random based on the proportion of the games that have that tag. The random predictions were made 1,000 times, with the predictions made independently at random, and the error rates were averaged.

\section{Experimental results}
\subsection{SVM Based Structured Classifier}
We built a support vector machine for structured classification of games. We filtered our tagged games by the number of plies (number of individual moves), generated mappings between tags and game prefixes (some initial position plus a sequence of moves), segmented our tagged games into three groups (training, validation and test), and called SVMLight with that data, an external library for SVM.

@@

\begin{table}
\begin{tabular}{lr}
\hline
\textbf{Tag} & \textbf{Error rate} \\ \hline
\tt{queenCapture}  & 0.006452 \\
\tt{kingInCenter}  & 0.006452 \\
\tt{perpetualCheck}  & 0.006452 \\
\tt{drawCombinationsStalemate}  & 0.038710 \\
\tt{bishopPair}  & 0.006452 \\
\tt{doubleAttack}  & 0.006452 \\
\tt{clearance}  & 0.025806 \\
\tt{openingClosingDiagonals}  & 0.006452 \\
\tt{openingClosingFiles}  & 0.006452 \\
\tt{enclosedKings}  & 0.012903 \\
\tt{discoveredAttack}  & 0.025806 \\
\tt{passedPawn}  & 0.012903 \\
\tt{quietMates}  & 0.038710 \\
\tt{pin}  & 0.051613 \\
\tt{removalOfDefence}  & 0.225806 \\
\tt{deflection}  & 0.251613 \\
\tt{decoy}  & 0.096774 \\
\tt{interference}  & 0.038710 \\
\tt{backrankWeakness}  & 0.032258 \\
\tt{freeingBlockingSquares}  & 0.019355 \\
\tt{pawnPromotion}  & 0.064516 \\
\tt{pullingPiecesToBadSquares}  & 0.006452
\end{tabular} 
\caption{False negative and false positive rates for various tags.}
\label{table:negatives}
\end{table}

@@

@@

\subsection{Classification Quality}
Future work should emphasize the classification quality. We demonstrated that general approaches may not be effective enough to be useful for complex games like chess. Therefore, future projects should seek improvements by better taking advantage of the chess-specific tools and libraries that are available like game engines and weight functions. If the classifier incorporated chess-specific features like those generated by chess engines, it could be much better at identifying the types of strategies that are available to the player. For example, a deep traversal of many future paths by a chess engine may reveal that an attack is possible (attackPossible) or that the player can force a pawn promotion (pawnPromotion). While a purely computational approach to identifying opportunities and strategies may provide higher-quality or more reliable assessments, a statistical approach may be more robust. %talk about what i mean by robust

@@

Another possible improvement is to use a different classifier than SVM, such as the perceptron or a maximum entropy classifier. One issue with the SVM is that often the tags are not so clear-cut: what does it mean that an endgame has “good chances”? Rather, there are degrees of how good an endgame is to the player. Therefore, a softer approach like maximum entropy may produce better results.

\subsection{Generalization to Other Games}
One goal of this project was to develop a statistical machine learning approach that was general enough to be easily adapted to other applications like StarCraft or Checkers. A future work can explore this possibility by simply creating features for other applications.
